---
title: "Assignment 2 Task 1 - Florida Palmetto Exploration"
author: "Katherine Rosencrance"
date: "2023-02-12"
output:
  html_document: 
    code_folding: hide
    theme: cerulean
---

# Overview

This report explores several characteristics of two different palmetto species at Archbold Biological Station in south-central Florida. Annual data measures were height, canopy length and width (all in cm), number of new and green leaves, and flowering scapes. Data were collected from 1981 through 1997 then again in 2001 and 2017 (Abrahamson, 2019). A binary logistic regression was performed to determine the probability of a plant being either *Serenoa repens* or *Sabal etonia* based on several predictor variables in two different models. The results evaluate how successfully the best model would “classify” a plant as the correct species.

# Setup

```{r setup, echo = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Attach packages
library(tidyverse)
library(here)
library(GGally)
library(broom)
library(AICcmodavg)
library(kableExtra)
library(tidymodels)

```

```{r}
# Read in the data
palmetto <- read_csv(here("data", "palmetto.csv"))
```

# Data Visualization and Analysis

Exploring differences in height, canopy length, canopy width, and green leaves for the two species.

```{r}
# Rename the 1 and 2 species codes to the actual species names
palmetto <- palmetto %>% 
   mutate(species = case_when(species == "1" ~ "Serenoa repens",
    species == "2" ~ "Sabal etonia"))

# Create a subset with the data we need
 palmetto_subset <- palmetto %>% 
   select(species, height:green_lvs) %>% 
   drop_na()
 
# Convert species to a factor, check to make sure it worked
palmetto_subset$species <- as.factor(palmetto_subset$species)
#class(palmetto_subset$species)
#levels(palmetto_subset$species)

# Note that Sabel etonia (0) is before Serenoa repens (1)

```

## Checking out trends across variables for the two species.

```{r}

# use ggpairs to view variables
# ggpairs(palmetto_subset, aes(color = species))


# Detailed graphs for some variables
height_length_plot <- ggplot(data = palmetto_subset, aes(x = height, y = length)) +
  geom_point(aes(color = species), show.legend = FALSE) +
  facet_wrap(~species)+
  theme_minimal()+
  labs( x = "Height (cm)",
        y = "Canopy Length (cm)",
        title = "Height and Canopy Length of Palmetto Species",
        subtitle = "1981-2017")+
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 11),
        axis.title.x = element_text(vjust = 0.2, face = "bold", size = 10),
        axis.title.y = element_text(face = "bold", size = 10),
        axis.text = element_text(size = 8),
        plot.background = element_rect(fill = "white"))+
  scale_color_manual(values = c("darkgreen", "lightgreen"))

  height_length_plot
```

**Figure 1: Height vs. Canopy Length of Palmetto Species.** Height and canopy length in centimeters of two different Palmetto species, *Sabel etonia* (dark green) and *Serenoa repens* (light green).

```{r}
height_width_plot <- ggplot(data = palmetto_subset, aes(x = height, y = width)) +
  geom_point(aes(color = species), show.legend = FALSE) +
  facet_wrap(~species)+
  theme_minimal()+
  labs( x = "Height (cm)",
        y = "Canopy Width (cm)",
        title = "Height and Canopy Width of Palmetto Species",
        subtitle = "1981-2017")+
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 11),
        axis.title.x = element_text(vjust = 0.2, face = "bold", size = 10),
        axis.title.y = element_text(face = "bold", size = 10),
        axis.text = element_text(size = 8),
        plot.background = element_rect(fill = "white"))+
  scale_color_manual(values = c("darkgreen", "lightgreen"))

height_width_plot
```

**Figure 2: Height vs. Canopy Width of Palmetto Species.** Height and canopy width in centimeters of two different Palmetto species, *Sabel etonia* (dark green) and *Serenoa repens* (light green).

```{r}
height_lvs_plot <- ggplot(data = palmetto_subset, aes(x = green_lvs, fill = species)) +
  geom_histogram() +
  facet_wrap(~species)+
  theme_minimal()+
  labs( x = "Number of Green Leaves",
        y = "Count per species",
        title = "Count of Green Leaves Across Palmetto Species",
        subtitle = "1981-2017")+
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
        plot.subtitle = element_text(hjust = 0.5, face = "bold", size = 11),
        axis.title.x = element_text(vjust = 0.2, face = "bold", size = 10),
        axis.title.y = element_text(face = "bold", size = 10),
        axis.text = element_text(size = 8),
        plot.background = element_rect(fill = "white"),
        legend.position = "none")+
  scale_fill_manual(values = c("darkgreen", "lightgreen"))

height_lvs_plot
```

**Figure 3: Number of Plants with Green Leaf Counts.** A histogram displaying the count of plants that have varying numbers of green leaves of two different Palmetto species, *Sabel etonia* (dark green) and *Serenoa repens* (light green).

Based on the plots above, and the visible difference between the two species, one could see that the count of green leaves would have the highest likelihood of classifying species correctly. Canopy length, width, and height are all fairly similar across the two species, however, length may be slightly better in predicting species.

# Binary Logistic Regression

```{r}
# Log odds of plant type using plant height, canopy length, canopy width and green leaves as predictor variable.
f1 <- species ~ height + length + width + green_lvs

palmetto_blr1 <- glm(formula = f1,
                    data = palmetto_subset,
                    family = "binomial")

# Log odds of plant type using plant height, canopy width and green leaves
f2 <- species ~ height + width + green_lvs

palmetto_blr2 <- glm(formula = f2,
                    data = palmetto_subset,
                    family = "binomial")

tidy(palmetto_blr1) %>% 
  kable(col.names = c("Term", "Estimate", "Standard Error", "Statistic", "P-value"),
        caption = "Table 1: Binary Logistic Regression Results Model 1") %>% 
  kable_classic(position = "center")

tidy(palmetto_blr2) %>% 
  kable(col.names = c("Term", "Estimate", "Standard Error", "Statistic", "P-value"),
        caption = "Table 2: Binary Logistic Regression Results Model 2") %>% 
  kable_classic(position = "center")
```
## AIC and BIC
```{r}
# AIC and BIC
aictab(list(palmetto_blr1, palmetto_blr2)) %>% 
    kable(col.names = c("Model", "Parameters", "AIC", "Delta AIC", "Model Likelihood", "AIC Weight", "Log Likelihood", "Cumulative Weight"),
        caption = "Table 3: AIC Results") %>% 
  kable_classic(position = "center")

bictab(list(palmetto_blr1, palmetto_blr2)) %>% 
    kable(col.names = c("Model", "Parameters", "BIC", "Delta BIC", "Model Likelihood", "BIC Weight", "Log Likelihood", "Cumulative Weight"),
        caption = "Table 4: BIC Results") %>% 
  kable_classic(position = "center")
```
## 10-fold Cross Validation
```{r}
# 10 fold cross validation
## set seed for reproducibility!
set.seed(345)

# set folds
tidy_folds <- vfold_cv(palmetto_subset, v = 10)
#tidy_folds

## use a workflow that bundles the logistic model and a formula
 blr_model <- logistic_reg() %>%
   set_engine('glm')

blr_tidy_wf1 <- workflow() %>%
  add_model(blr_model) %>%
  add_formula(f1)

blr_tidy_cv_f1 <- blr_tidy_wf1 %>%
  fit_resamples(tidy_folds)

## use functions from the tune package to extract metrics
metrics1<- collect_metrics(blr_tidy_cv_f1)

# make a nice table to display model 1 metrics from CV
nice_metrics<- metrics1 %>% 
  kable(col.names = c("Metric", "Estimator", "Mean", "n", "Standard Error", "Configuration"),
        caption = "Table 5: Cross Validation Metrics Model 1") %>% 
  kable_classic(position = "center")

nice_metrics

## mean accuracy is 0.9168 for model 1


## Repeat for model 2 
blr_tidy_wf2 <- workflow() %>%
  add_model(blr_model) %>%
  add_formula(f2)

blr_tidy_cv_f2 <- blr_tidy_wf2 %>%
  fit_resamples(tidy_folds)

## use functions from the tune package to extract metrics
metrics2<- collect_metrics(blr_tidy_cv_f2)

nice_metrics_2<- metrics2 %>% 
  kable(col.names = c("Metric", "Estimator", "Mean", "n", "Standard Error", "Configuration"),
        caption = "Table 6: Cross Validation Metrics Model 2") %>% 
  kable_classic(position = "center")

nice_metrics_2

## mean accuracy is 0.8988 for model 2

```

Based on the AIC and BIC values, as well as the 10-fold cross validation, model 1 performs better. The AIC and BIC for model 1 were lower and the mean accuracy metric was `r round(metrics1[1,3], 5)` compared to `r round(metrics2[1,3], 5)` for model 2.

```{r}
# use the entire dataset, rather than testing/training sets, to identify the coefficients for the final predictive model, based on model 1 (lowest AIC and BIC)
blr_tidyfit_f1 <- blr_model %>% 
  fit(f1, data = palmetto_subset)

coefficient_df <- blr_tidyfit_f1 %>% 
  tidy() %>% 
  kable(col.names = c("Term", "Estimate","Standard Error", "Statistic", "P-Value"),
        caption = "Table 7: Coefficients for Final Predictive Model") %>% 
  kable_classic(position = "center")

coefficient_df

```

```{r}
# But log odds are challenging to interpret. Let's find actual *probabilities* associated with a palmetto being S etonia or S repens, based on model 1
blr1_fitted <- palmetto_blr1 %>%
  broom::augment(type.predict = "response") %>% 
  
# add columns to see if the model correctly guessed the species
  
  mutate(species_predicted= case_when(.fitted >= 0.5 ~ "Serenoa repens",
         TRUE ~ "Sabal etonia")) %>% 
  mutate(correct_prediction= case_when(species == species_predicted ~ "yes",
                                        TRUE ~ "no"))

#create table
palmetto_table <- blr1_fitted %>% 
  select(species, correct_prediction) %>% 
  pivot_wider(names_from = correct_prediction, values_from = correct_prediction, values_fn =   list(correct_prediction = length)) %>% 
  mutate(percent_correctly_classified= yes/(yes+no)*100)

nice_palmetto_table<- palmetto_table %>% 
 kable(col.names = c("Species", "Correctly Classified", "Incorrectly Classified", "% Correctly Classified"),
       caption = "Table 7: Classification Results using Model 1") %>% 
 kable_classic(position = "center")

nice_palmetto_table

```

# Conclusion

Model 1 (which included the variables plant height, canopy length, canopy width and green leaves) was proven to statistically predict species better than model 2 (which did not include canopy length). This was supported through AIC results, BIC results, and 10-fold cross validation. As shown in Table 3, the percent correctly classified as *Serenoa repens* was `r palmetto_table[1,4]` % and the percent for *Sabel etonia* was `r palmetto_table[2,4]` %. These numbers are fairly high, additionally supporting the choice of the model 1 predictor variables.

# Citation

**Data Citation:** Abrahamson, W.G. 2019. Survival, growth and biomass estimates of two dominant palmetto species of south-central Florida from 1981 - 2017, ongoing at 5-year intervals ver 1. Environmental Data Initiative. <https://doi.org/10.6073/pasta/f2f96ec76fbbd4b9db431c79a770c4d5>
